---
title: "NOAA's Storm Events Database:\n A Swiss knife in preparing for severe weather events"
output: 
  html_document:
    keep_md: true
---

[Rpubs link] (http://rpubs.com/mkoniari/59800)

## Synopsis

In this report, we aim to analyze the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We use the number of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy. From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

## Loading and preprocessing the raw data

Data for this assignment can be downloaded from:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]
There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Reading the raw data

There are 902297 rows and 37 columns in total. The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete


```{r preload, message=FALSE, warning=FALSE}
remove(list=ls())
sessionInfo()
# load r libraries
packages <- c("knitr", "dplyr", "tidyr",  "ggplot2", "gridExtra", "R.utils", "downloader", "lubridate")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}
sapply(packages, require, character.only=TRUE, quietly=TRUE)


downloadFile <-function(furl, fsave){
        download(furl, fsave = "wb")
        #download.file(fileUrl,  fsave, method="curl")
        cat(paste0("file download complete - remember to unzip file: ",fsave))        
}   

#DEBUG <- TRUE
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
DATAFILE <- file.path(getwd(), "stormData.cvs")
DATAZIP  <- file.path(getwd(), "stormData.bz2")

# check and download data file
if (!file.exists(DATAFILE)) {
  cat(paste0(DATAFILE," doesn't exist"))
  # check data exists
        if (!file.exists(DATAZIP)) {
                cat(paste0(DATAZIP," doesn't exist trying to download ... "))
                downloadFile(fileUrl, DATAZIP)
                }
  # Extract file from bz2
  bunzip2(DATAZIP, DATAFILE, remove = FALSE)  
  stopifnot(file.exists(DATAFILE))          
}   
```

```{r read_data, cache=TRUE}
assign( "rawData", read.csv(DATAFILE), envir = .GlobalEnv)
```

### Standardizing EVTYPEs

Data is rather messy. There are 985 factors of EVTYPE. So we need to clean the EVTYPE label.  We'll collapse the many variations of the most common event types into a single EVTYPE.  In general we do  `head(sort(table(rawData$EVTYPE), decreasing=TRUE), 30)`
and observe high frequency counts that could be combined, e.g. 
`table(grep(“THUNDERSTORM WIND”, rawData$EVTYPE, value=TRUE))`
We use the EVTYPE with highest frequency as the “standard” name.

```{r EVTYPE_fix}
head(table(rawData$EVTYPE), n =20)

# Force all EVTYPEs to uppercase to avoid case variations
rawData$EVTYPE <- toupper(rawData$EVTYPE)

# Standardize spacing: replace multiple spaces with single space
rawData$EVTYPE <- gsub(" +", " ", rawData$EVTYPE)

rawData$EVTYPE[grep("HURRICANE|TYPHOON|TROPICAL", rawData$EVTYPE)] <- "HURRICANE/TYPHOON/TROPICAL STORM"

rawData$EVTYPE[grep("SNOW|ICE|BLIZZARD|WINTER STORM|FREEZING|WINTRY MIX|SLEET", rawData$EVTYPE)] <- "WINTER STORM"

rawData$EVTYPE[grep("WINTER WEATHER|COLD|ICY|GLAZE|LOW TEMPERATURE|HYPOTHERMIA|HYPERTHERMIA|COLD TEMPERATURE|WINDCHILL", 
    rawData$EVTYPE)] <- "WINTER WEATHER"

# ~15 variations of 'High Wind' and 3 variations of 'strong wind' but after
# 'Snow' variations removed above
rawData$EVTYPE[grep("HIGH WIND|STRONG WIND|Gusty Winds|GUSTY WIND", rawData$EVTYPE)] <- "HIGH WIND"

# Change these 'false hits' for 'thunderstorm wind' to wind
rawData$EVTYPE <- sub("NON-THUNDERSTORM WIND|NON THUNDERSTORM WIND", "WIND", rawData$EVTYPE)

# Other non thunderstorm winds

# Standardize variations/spelling errors in 'Thunderstorm'
rawData$EVTYPE <- gsub("THUNDEERSTORM|THUNDERESTORM|THUNDERSTROM|THUDERSTORM|THUNERSTORM|THUNDERTORM|TUNDERSTORM", 
    "THUNDERSTORM", rawData$EVTYPE)

rawData$EVTYPE <- sub("TSTM WIND|STORM FORCE WINDS", "THUNDERSTORM WIND", rawData$EVTYPE)
# Manually reviewed ~44 variations that are combined here
rawData$EVTYPE[grep("THUNDERSTORMW|THUNDERSTORMWINDS|THUNDERSTORM WINS|TSTMW", rawData$EVTYPE)] <- "THUNDERSTORM WIND"
rawData$EVTYPE[grep("THUNDERSTORM WIND|THUNDERSTORMS WIND", rawData$EVTYPE)] <- "THUNDERSTORM WIND"

# Combine ~59 variations of FLOODING
rawData$EVTYPE[grep("FLOOD|FLD|RECORD RAINFALL", rawData$EVTYPE)] <- "FLOODING"

# Combine ~20 variations of HAIL
rawData$EVTYPE[grep("HAIL", rawData$EVTYPE)] <- "HAIL"

# Combine variations of LIGHTNING
rawData$EVTYPE[grep("LIGHTNING|LIGHTING|LIGNTNING", rawData$EVTYPE)] <- "LIGHTNING"

# Combine variations of fire
rawData$EVTYPE[grep("HEAVY SURF|HIGH SURF", rawData$EVTYPE)] <- "HEAVY/HIGH SURF"

# Combine variations of fire
rawData$EVTYPE[grep("FIRE", rawData$EVTYPE)] <- "WILDFIRE"

# Combine variations of freeze -- likely agriculture related
rawData$EVTYPE[grep("FROST|FREEZE", rawData$EVTYPE)] <- "FROST/FREEZE"

rawData$EVTYPE[grep("SLIDE", rawData$EVTYPE)] <- "LAND/MUD/ROCK SLIDES"

rawData$EVTYPE[grep("RIP", rawData$EVTYPE)] <- "RIP CURRENT"

# HEAT or EXCESSIVE HEAT
rawData$EVTYPE[grep("HEAT", rawData$EVTYPE)] <- "HEAT"

rawData$EVTYPE[grep("DENSE FOG", rawData$EVTYPE)] <- "FOG"

# Only notable winds are high wind
rawData$EVTYPE[rawData$EVTYPE == "WINDS" | rawData$EVTYPE == "WIND"] <- "HIGH WIND"


# Collapse all events with frequency < 50 to category “mis”:

event.types <- sort(table(rawData$EVTYPE), decreasing = TRUE)
event.types.under.50 <- names(event.types[event.types < 50])

rawData$EVTYPE[rawData$EVTYPE %in% event.types.under.50] <- "MISC"
```

Thus the list of EVTYPEs for the analysis:
```{r EVTYPE_list}
sort(table(rawData$EVTYPE), decreasing = TRUE)

```

We keep only the records that show fatalities, injuries, property or crop damage.

```{r check_data}
stormData <- tbl_df(rawData) %>%
        filter(FATALITIES >0 | INJURIES >0 | PROPDMG >0| CROPDMG >0) %>%
        mutate (BGN_DATE  = mdy_hms(BGN_DATE), year = year(BGN_DATE))


colSums(is.na(stormData))
summary(stormData)
```
There are 902297 rows and 37 columns in total. From them we kept only 254633 interesting observations. 

#### A note on the temporal aspect of measures 

```{r temp}
temporal <- stormData %>% 
        mutate(year= year(BGN_DATE)) %>%
        filter(year<1990) %>%
        group_by(year) %>%                              
        summarise(Events  = n(),
                Unique_Elements = n_distinct(EVTYPE)) 
```
We have to note that data defore year 1990 has only **`r max(temporal$Unique_Elements)` types ** of observations. This finding raises questions as to whether we should ignore early years data (before 1990) or not. We choose to include the data in our study.


### Fix property and crop damage data to be useful in analysis
We will convert the **property damage** and **crop damage** data into comparable numerical forms according to the meaning of units described in the code book ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)). Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).


```{r damageFix}
multiplierAbr <- c("H", "K", "M", "B")
multiplier <- c(100, 1000, 1e+06, 1e+09)
names(multiplier) <- multiplierAbr
# multiplier['H'] == 100

stormData <- stormData %>% 
        mutate(PROPDMGEXP = toupper(PROPDMGEXP), CROPDMGEXP = toupper(CROPDMGEXP)) %>%
        mutate(propCash = ifelse(PROPDMGEXP %in% multiplierAbr, multiplier[PROPDMGEXP]*PROPDMG,0.0)) %>%
        mutate(cropCash = ifelse(CROPDMGEXP %in% multiplierAbr, multiplier[CROPDMGEXP]*CROPDMG,0.0))
```

## Results

        1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?          
We check the number of fatalities and injuries that are caused by the severe weather events. We would like to get the most severe types of weather events.        

```{r fatalStorm}
healthStorm <- stormData %>% 
                group_by(EVTYPE) %>%                              
                summarise(
                        Events  = n(), 
                        Fatalities = sum(FATALITIES), 
                        Injuries = sum(INJURIES)) %>%
                arrange(desc(Fatalities), desc(Injuries)) %>%
                print



top5fatalEvents <- healthStorm  %>% 
                        arrange(desc(Fatalities)) %>% 
                        select ( EVTYPE,Fatalities) %>% 
                        mutate(EVTYPE = reorder(EVTYPE, Fatalities, max)) %>%
                        slice(1:5)

top5injEvents <- healthStorm  %>% 
                        arrange(desc(Injuries)) %>% 
                        select (EVTYPE,Injuries) %>% 
                        mutate(EVTYPE = reorder(EVTYPE, Injuries, max)) %>%
                        slice(1:5)
```

The most harmful with respect to Fatalities in population health are:**`r top5fatalEvents$EVTYPE` **.

The most harmful with respect to Injuries in population health are: **`r top5injEvents$EVTYPE` **.

We plot total fatalities and total injuries. 

```{r fatalPlot}
# plot fatalities
fplot <- ggplot(data=top5fatalEvents, aes(EVTYPE, Fatalities)) + 
    geom_bar(colour="black",  width=.7, stat="identity") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Severe Weather Type") + ylab("Fatalities") +
    ggtitle("Fatalities by Severe Weather\n Events in the U.S.")
# plot Injuries
iplot <- ggplot(data=top5injEvents, aes(EVTYPE, Injuries)) + 
    geom_bar(colour="black",  width=.7, stat="identity") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Severe Weather Type") + ylab("Injuries") +
    ggtitle("Injuries by Severe Weather\n Events in the U.S.")
grid.arrange(fplot, iplot, ncol = 2)
```
        2. Across the United States, which types of events have the greatest economic consequences?
```{r economyPlot}
ecoStrom <- stormData %>% 
                group_by(EVTYPE) %>%                              
                summarise(
                        Events  = n(), 
                        propertyDamage = sum(propCash) / 10^9, 
                        cropDamage = sum(cropCash)/10^9) %>%
                arrange(desc(propertyDamage), desc(cropDamage)) %>%
                print

top5PropEvents <- ecoStrom  %>% 
                        arrange(desc(propertyDamage)) %>% 
                        select ( EVTYPE,propertyDamage) %>% 
                        mutate(EVTYPE = reorder(EVTYPE, propertyDamage, max)) %>%
                        slice(1:5)

top5CropEvents <- ecoStrom  %>% 
                        arrange(desc(cropDamage)) %>% 
                        select (EVTYPE,cropDamage) %>% 
                        mutate(EVTYPE = reorder(EVTYPE, cropDamage, max)) %>%
                        slice(1:5)
```

The events that have the greatest economic consequences are: **`r top5PropEvents$EVTYPE` **. 

The most harmful events with respect to crop production are: **`r top5CropEvents$EVTYPE` **.

We then plot total property damage and total crop damage. 

```{r ecoPlot}
# plot propertyDamage
pplot <- ggplot(data=top5PropEvents, aes(EVTYPE, propertyDamage)) + 
    geom_bar(colour="black",  width=.7, stat="identity") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Severe Weather Type") + ylab("Property Damage") +
    ggtitle("Property Damage by Severe Weather\n Events in the U.S. (Billion $)")
# plot cropDamage
cplot <- ggplot(data=top5CropEvents, aes(EVTYPE, cropDamage)) + 
    geom_bar(colour="black",  width=.7, stat="identity") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Severe Weather Type") + ylab("Crop damage") +
    ggtitle("Crop damage by Severe Weather\n Events in the U.S. (Billion $)")
grid.arrange(pplot, cplot, ncol = 2)
```
